Brief Announcement: Tight Bounds for Online Balanced Repartitioning

### 2020-05-18 podc_NEW/podc_ba.tex
Cutting a lot

This problem is known as the dynamic balanced graph partitioning problem and was introduced by Avin et al.~\cite{repartition-disc, sidma-arxiv} at DISC 2016. a~special variant of the general problem has later been studied by Henzinger et al.~\cite{sigmetrics19_partitioning} at SIGMETRICS 2019.
We refer to the latter as the learning model.

\subsection{Model}

We study two models in this paper: the \emph{general partitioning} model, and its subproblem, the \emph{learning} model.
In both models, we assume that communication patterns are not known to our algorithms at the beginning.
We measure the~quality of~presented algorithmic solutions by competitive analysis~\cite{borodin-book}, which is well-suited for problems that are online by their nature.
In the competitive analysis, the goal is to~optimize \emph{the competitive ratio} of a~given online algorithm: the ratio of its cost to the cost of~an~optimal offline algorithm that knows the entire input sequence in advance.
%We emphasize that all our algorithms have a~\emph{strict} competitive ratio (without an additive term).

\noindent
\textbf{General partitioning model.}
In the \emph{dynamic balanced graph partitioning} problem, we are given a~set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a~sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a~pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

@@

\noindent
\textbf{Learning model.}
We study a~\emph{learning} variant of dynamic balanced graph partitioning,
where the communication pattern is \emph{static}:
whether a~pair of  nodes ever communicate or not, 
is determined a~priori and is unknown to algorithms,
 and such pairs communicate forever.
Any algorithm must eventually collocate  pairs of communicating nodes,
as otherwise it cannot be competitive.
As in Henzinger et al.~\cite{sigmetrics19_partitioning}, we assume that the communication graph admits a~\emph{perfect partition},
i.e., a~partition in which no inter-cluster request ever occurs.
%any algorithm must learn this perfect partition.
The algorithm's objective is to \emph{learn} the (static) communication graph
 while serving all requests,
and without executing too many migrations.
For the learning model, we  assume that the migration cost is $\alpha=1$.

\subsection{Related Work}

The two works closest to ours are by Avin et al. (on the general partitioning model)~\cite{repartition-disc, sidma-arxiv} and by Henzinger et al. (on the learning model)~\cite{sigmetrics19_partitioning}.
However, the focus of these papers is primarily on models with resource augmentation: the online algorithm can use slightly larger clusters than the offline algorithm.  
Avin et al. actually showed that their lower bound $\Omega(k)$ holds even for a~significant resource augmentation, and they provided an algorithm with the competitive ratio $O(k \log k)$ using the $(2+\epsilon)$-augmented cluster capacity.
Their ratio is independent of $\ell$, which is impossible without significant resource augmentation.

In contrast, we study the non-augmented setting, where the nodes need to be perfectly balanced  among the clusters.
This assumption is not only more realistic but also significantly more challenging, as it is related to hard problems such as integer partitioning~\cite{integer-partitions-book}.
In terms of results without augmentation, so far, it is only known that there exists an~$O(k^2 \cdot \ell^2)$-competitive algorithm~\cite{repartition-disc}; the best known lower bound is significantly lower, namely $\Omega(k)$.
For $k=2$, Avin et al.~\cite{repartition-disc} presented a~$7$-competitive algorithm with a~substantial ($\Omega(\ell^2)$) additive constant.

The problem has also been studied in a~weaker
model where the adversary can only sample
requests from a~fixed distribution~\cite{stochastic-ring}.

The static offline version of~the~partitioning~problem, i.e., a~problem variant where
migration is not allowed, where all requests are known in advance, and where
the goal is to find an assignment of $n$ nodes to $\ell$~physical machines, each of~capacity $n/\ell$, is known as the
\emph{$\ell$-balanced graph partitioning problem}. The problem is 
NP-complete, and cannot even be approximated within any finite factor unless P
= NP~\cite{AndRae06}.  The static
variant where $\ell = 2$ corresponds to the minimum bisection problem, which
is already NP-hard~\cite{GaJoSt76}, and 
the currently best approximation ratio is $O(\log n)$~\cite{SarVaz95,ArKaKa99,FeKrNi00,FeiKra02,KraFei06,Raec08}.

Our problem is further related to some classic online problems.
In particular, it is related to online paging~\cite{SleTar85,FKLMSY91,McGSle91,AcChNo00}, sometimes also referred to
as online caching, where requests for data items (nodes) arrive over time and
need to be served from a~cache of finite capacity, and where the number of
cache misses must be minimized. Classic problem variants usually boil down to
finding a~smart eviction strategy, such as Least Recently Used (LRU)~\cite{SleTar85}. In our
setting, requests can be served remotely (i.e.,~without fetching the
corresponding nodes to a~single physical machine). In this light, our model is more
reminiscent of caching models \emph{with
bypassing}~\cite{EpImLN11,EpImLN15,Irani02}.
A major difference between  these problems is that in the caching problems, each request involves a~single element of the universe, while in our model \emph{both} endpoints of a~communication request are subject to~optimization.
In this light, we can see our model as a~"symmetric" version of online paging.

Dynamic graph partitioning problems are generally fundamental in computer science, and arise in many different contexts~\cite{streaming-soda,streaming1}.

%More generally, the model is related to online
%caching~\cite{SleTar85,FKLMSY91,McGSle91,AcChNo00},
%see~\cite{repartition-disc} for a~discussion.
%The static offline version of~the~problem, called the
%\emph{$\ell$-balanced graph partitioning problem} is 
%NP-complete, and cannot even be approximated within %any finite factor unless P
%= NP~\cite{AndRae06}.

\subsection{Our Contributions}

This paper presents several new results on the dynamic graph partitioning problem  without augmentation.
For the learning model, we present a~lower bound of $\Omega(k\cdot\ell)$ on the competitive ratio of any online deterministic online algorithm 
(that holds also in the general partitioning model).
The best known lower bound so far was $\Omega(k)$~\cite{repartition-disc} that holds only in the general partitioning model.
We complement this result with an asymptotically optimal, 
$O(k\cdot \ell)$-competitive algorithm
for the learning model.

For the general partitioning model, we design
an asymptotically optimal,
$\Theta(\ell)$-competitive algorithm for $k=3$, improving the best known upper bound 
so far $O(\ell^2)$~\cite{repartition-disc}.
We further present a~strictly $6$-competitive algorithm for $k=2$ that improves upon the previous $7$-competitive algorithm with $O(\alpha\ell^2)$ additive constant.

All algorithms in this paper have a strict competitive ratio (i.e., without an additive term).

\medskip
\noindent
\textbf{Algorithmic techniques.} The variant for general $k$ is still unresolved, hence it is vital to summarize algorithmic techniques used in this paper.
A straightforward analysis of the algorithm from Section~\ref{sec:k3} results in the bound $O(\ell^2)$, and the analysis must be improved in two places. One must bound the cost of each of $O(\ell)$ reconfigurations per~phase by a~constant, and show that the optimum offline algorithm must pay a~significant cost for inter-cluster requests.
We bound the cost of the latter by estimating the capabilities of the optimum offline algorithm to \emph{prepare} for an incoming sequence of requests.
Furthermore, in the analysis of the algorithm from Section~\ref{sec:k2}, we propose a~novel charging scheme for edges that share a~vertex.
We also emphasize the implications of the tight result for the learning model.
If one hopes to improve the lower bound for the general model, the ground sets must be dynamically added and \emph{removed}: if ground sets are only added, the instance is solvable by the algorithm from Section~\ref{sec:ppl} within the cost $O(k\cdot \ell)$.

>>>>

\noindent \textbf{Model.}
The \emph{dynamic balanced graph partitioning} problem (\OBRP{})
is a fundamental learning problem
that finds applications in the context of
distributed systems optimization~\cite{repartition-disc}.
We are given a~set $V$ of $n$ nodes 
(e.g., virtual machines or processes),
initially arbitrarily partitioned into $\ell$~clusters
(e.g., servers or entire racks),
each of size~$k$.
The nodes interact using
a~sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a~pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a~certain amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.

@@

We consider two flavors of the problem
in this paper. In the first one, which
was also studied by Henzinger et al.~\cite{sigmetrics19_partitioning}
(SIGMETRICS 2019) and which
we here mainly consider for the sake of a stronger
lower bound, we assume that $\sigma$
simply reveals the edges of a static graph
that can be perfectly partitioned
(i.e., in principle, no external requests
are required). 
In the second one, we consider a general
model where $\sigma$ can be arbitrary;
this was the original problem introduced
by Avin et al.~\cite{repartition-disc} at DISC 2016.
For simplicity, we will refer to the former
model as the \emph{learning} model (as
one has to learn the static communication graph) 
and to the latter as the \emph{online} model.

\noindent \textbf{Related work.}
%
The two works closest to ours are by Avin et al. (on the general partitioning model)~\cite{repartition-disc, sidma-arxiv} and by Henzinger et al. (on the learning model)~\cite{sigmetrics19_partitioning}.
The static offline version of~the~partitioning~problem, i.e., a~problem variant where
migration is not allowed, where all requests are known in advance, and where
the goal is to find an assignment of $n$ nodes to $\ell$~physical machines, each of~capacity $n/\ell$, is known as the
\emph{$\ell$-balanced graph partitioning problem}~\cite{AndRae06}.
Dynamic graph partitioning problems are generally fundamental in computer science, and arise in many different contexts~\cite{streaming-soda,streaming1}.

\noindent \textbf{Contributions.}
%
This paper presents several new results on the dynamic graph partitioning problem  without augmentation.
For the learning model, we present a~lower bound of $\Omega(k\cdot\ell)$ on the competitive ratio of any online deterministic online algorithm 
(that holds also in the general partitioning model).
The best known lower bound so far was $\Omega(k)$~\cite{repartition-disc} that holds only in the general partitioning model.
We complement this result with an asymptotically optimal, 
$O(k\cdot \ell)$-competitive algorithm
for the learning model.
For the general partitioning model, we design
an asymptotically optimal,
$\Theta(\ell)$-competitive algorithm for $k=3$, improving the best known upper bound 
so far $O(\ell^2)$~\cite{repartition-disc}.
We further present a~strictly $6$-competitive algorithm for $k=2$ that improves upon the previous $7$-competitive algorithm with $O(\alpha\ell^2)$ additive constant.
